Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-09-12 main.tex
Made a background section separate from the more general introduction.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large number of
reported injuries~\citep{Wegman2024}. At low speeds, from start up to typical
cruising speeds, bicycles themselves are not self-stable and can be challenging
for the rider to balance. Low speed crashes may be reduced if the bicycle was
self-stable at these speeds by relieving the rider of a necessary control
activity. Bicycles can be mechanically modified to lower the speeds at which
they are self-stable~\citep{Astrom2005} and since the 1980s it has been known
that applying a motor actuated steering torque proportional to the roll rate
can stabilize a single track vehicle at very low speeds. If this automatic
control of steering can stabilize a bicycle, it may reduce the control required
by the rider to successfully manage balancing tasks. We have developed such a
balance assisting bicycle and hypothesize that it helps the rider in situations
in which they are likely to fall, Figure~\ref{fig:balance-assist-bicycle}.
\todo{Talk about the types of real life perturbations an how our perturbation
mimics it.}
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

In the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
attempts at automatic roll stabilization of single track vehicles began in
earnest after predictive motorcycle models were developed and refined in
throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
and seems to be the first to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities. Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, no human rider was involved. Many more automatically balanced single
track vehicles have been demonstrated over the last 40 years, but none have
demonstrated that increasing low speed stability can assist in balancing and
possibly reduce single actor falls. Most of these robotic bicycle and
motorcycle designers did not intend for a human rider to also control the
stabilized vehicle. Nevertheless, an automatically stabilized bicycle can be
controlled by a human rider if the motor controlled steer torque and the rider
applied steer torque act on the steer in parallel.  The effect of this
automatic control gives the ability to change the dynamics, up to some limits,
of the human-controlled plant. In our prior study~\citep{Alizadehsaravi2023},
we demonstrate reduced motion during distractions due to the balance assist but
contrarily researchers with a similar vehicle recently showed rider
dissatisfaction with the stabilization~\cite{Hanakam2023}.

@@

In this paper, we test whether a linearly gain scheduled steer motor controlled
bicycle, that is stable in a large low speed range when riderless, is
beneficial in helping to prevent the rider from falling. We test this by
applying varying magnitude mechanical perturbations to the handlebars while the
rider is cycling on a treadmill. We assess the difference in the rider's
probability of falling with the balance assist steer motor system on and off.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of the necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the roll rate can stabilize a single-track vehicle at
lower speeds than without the control. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks. We have developed such a balance assisting
bicycle, Figure~\ref{fig:balance-assist-bicycle}, and hypothesize that it helps
the rider in situations in which they are likely to fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handledbar, which can easily cause a rider to
fall, especially when the rider is not so attentive or has one or more hands
removed from the handlebar.

In this paper, we test whether a linearly gain scheduled steer motor controlled
bicycle, that is stable in a large low speed range, is beneficial in helping to
prevent the rider from falling. We test this by applying varying magnitude
mechanical perturbations to the handlebars while the rider is cycling on a
treadmill. We assess the difference in the rider's probability of falling with
the balance assist steer motor system on and off.

\subsection{Background}
%
In the early years of developments in automatic control, \citep{Whipple1899}
not only derived the correct equations of motion of the bicycle but realized
and showed that roll motion feedback can stabilize a bicycle. Much later,
attempts at automatic roll stabilization of single track vehicles began in
earnest after predictive motorcycle models were developed and refined in
throughout the 1970s. \Citet{vanZytveld1975} was influenced by Whipple's work
and seems to be the first to attempt to robotically stabilize a small motorbike
with a controlled inverted pendulum that mimicked rider lean, but he was not
successful in demonstrating what his theoretic control model correctly
predicted. In his model, he recognized that feedback of the vehicle roll angle
and angular rate was essential to stabilize the vehicle. It was not until the
early 1980s that \cite{Nagai1983} successfully demonstrated balancing a robotic
bicycle on a treadmill with both steering control and a laterally moving mass.
Ruijs and Pacjecka followed this by demonstrating an automatically balanced
motorcycle~\citep{Ruijs1986} and they did so with a steering motor. Ruijs and
Pacejka clearly showed that steer torque driven by roll angle feedback
stabilizes the capsize mode, roll angular rate feedback stabilizes the weave
mode, and steer angular rate feedback stabilizes the wobble mode. They also
proposed how the gains must change with respect to vehicle speed for favorable
control at all speeds. This roll motion feedback is the simplest controller
that can stabilize a single track vehicle above a minimum speed when one is not
concerned with wobble instabilities. Ruijs and Pacejka's work was not
particularly concerned with low speed stability and their vehicle was fully
automatic, no human rider was involved. Many more automatically balanced single
track vehicles have been demonstrated over the last 40 years, but none have
demonstrated that increasing low speed stability can assist in balancing and
possibly reduce single actor falls. Most of these robotic bicycle and
motorcycle designers did not intend for a human rider to also control the
stabilized vehicle. Nevertheless, an automatically stabilized bicycle can be
controlled by a human rider if the motor controlled steer torque and the rider
applied steer torque act on the steer in parallel.  The effect of this
automatic control gives the ability to change the dynamics, up to some limits,
of the human-controlled plant. In our prior study~\citep{Alizadehsaravi2023},
we demonstrate reduced motion during distractions due to the balance assist but
contrarily researchers with a similar vehicle recently showed rider
dissatisfaction with the stabilization~\cite{Hanakam2023}.

@@
